### 2013-12-30 mcmc.tex
estimators

Say you are in this situation.
You have a huge blob of data $\data$ (think of this as a vector or
list or heterogeneous collection of observations\note{In general in
  this \documentname\ we will treat any variable as being capable of
  containing not just a value but some huge arbitrarily structured set
  of values in matrix, vector, or more complex form.}).
You also have a model sophisticated enough---a probabilistic,
generative model, if you will\note{For our definition of a
  \emph{model}, see [SOME OTHER NOTE] in this series.}---that, given a
setting of a huge blob of parameters\note{Again, the parameter
  blob $\pars$ should also be thought of as an immense, arbitrarily
  structured set or list of parameters, possibly even with no fixed
  dimension!} $\pars$, you can compute a pdf for data (or likelihood\note{We
  bash the terminology ``likelihood'' in [SOME OTHER NOTE] in this
  series.}) $p(\data\given\pars)$.
% ************* BJB
% <Wild Subjective Bayesian>
% The "generative model" is a probabilistic description of your *prior
% beliefs about what the data are going to be*. </Wild Subjective Bayesian>
% ************* Hogg
% That comment should go in the ``What is a Model?'' Chapter.
Furthermore, say you also can write down some kind of informative or
vague prior pdf $p(\pars)$ for the parameter blob.
If all these things are true, then---even if you can't compute
anything else---in principle a trivial-to-implement MCMC can give you
a fair sampling of the posterior pdf.
That is, you can run MCMC (for a very long time\note{Sometimes a very,
  very long time; we will return to that below.})  and you will be
left with a set of $K$ parameter-blob settings $\pars_k$ such that the
full set $\setof{\pars_k}_{k=1}^K$ constitutes a fair sampling from
the posterior pdf $p(\pars\given\data$).
% ************* BJB
% Notions of "sampling" and "Monte Carlo" are interesting. You might like to
% refer the interested reader to the paper "Monte Carlo is Fundamentally
% Unsound" by Tony O'Hagan.

@@

Imagine that you are in the simplest case:
You have a model with a small number of parameters (say three-ish),
  and only one of them is of great interest.
What are your options for the ``measurement'' of this parameter?
The only simple integral-based options are the posterior mean
  or posterior median value for the parameter.
\\HOGG CHOOSE NOTATION AND GIVE FORMULAE HERE.

What are your best options for the ``uncertainty'' or ``limits'' on this parameter?
The only simple integral-based options are posterior quantiles.
That is, the ``one-sigma'' error bar can be the size of the central (or smallest%
  \note{HOGG shizzle about choosing quantiles}) interval of the parameter
  that contains 68 percent of the posterior samples.
\\HOGG GIVE FORMULA HERE.\\
The ``two-sigma'' error bar would be the same but for the 95-percent interval.

One amusing thing is that if you choose the 68-percent interval sensibly,
  but use as the ``measurement'' the posterior mean,
  you can get pathological situations in which the posterior mean measurement
  is actually \emph{outside} the one-sigma confidence interval.
For this reason (and others), we usually recommend as a default behavior%
  ---in the one-dimensional case---%
  to choose the \emph{median} of sampling as the measurement value,
  the 16-percent quantile as the lower error bar,
  and the 84-percent quantile as the upper error bar.
This has pathologies in higher dimensions (as we are about to see),
  but is pretty safe for one-dimensional answers.

@@

Another proxy for autocorrelation time is the
  Expected Squared Jump Distance.\note{(CITE GELMAN)}
This is the mean squared distance the walker moves, per step.
It is maximized when the acceptance ratio is reasonable
  and the step size is large;
  it is large when the exploration of the space is fast.
This proxy is easy to measure and use for tuning,
  and more directly related to autocorrelation time than the acceptance ratio.
We recommend using it for tuning, though we have never used it ourselves.
Like the acceptance fraction, it can also be used to tune
  in the case that the proposal distribution loops over parameters;
  once again, the user gets (in this case) $d$ one-dimensional tunings.

@@

MCMC is used to obtain samples $x_k$ from pdf $p(x)$,
  or a badly normalized all-positive function $f(x)$
  that is different from the pdf by an unknown factor $Z$.
In the context of data analysis,
  MCMC is usually being used to obtain samples
  that are \emph{parameter values} for a probabilistic model of some data.
That is, MCMC is sampling a pdf \emph{for the parameters}.
If it is returning parameter values, then it is not sampling a pdf for the data.
That is, MCMC can only sample a posterior (or prior) pdf for the parameters.
\emph{It cannot sample a likelihood.}%
\note{Well, technically, MCMC \emph{can} be used to sample a likelihood function,
  but the samples would be samples of possible \emph{data} not possible \emph{parameters}.
  The purist might say that even this is not sampling a likelihood function,
  because you should only call it a ``likelihood function'' in contexts in which you are
  treating the data as fixed (at the true data values) and the parameters as variable.}

@@

\paragraph{parameterization problems:}
It helps to think carefully about how to parameterize.
For example, if there is an angle $\phi$ in your problem,
  and your prior requires it to be between $0$ and $2\pi$,
  and the posterior mode is very near $0$,
  a little bit of probability leaking below $0$ plus the prior cutoffs plus mod-$2\pi$ symmetry
  makes an intrinsically unimodal problem multimodal!
In these cases, we usually advise reparameterization from a (say) amplitude $A$ and angle $\phi$
  to two vector components $A\,\cos\phi$ and $A\,\sin\phi$.
Of course this re-parameterization changes significantly the \emph{form} of the prior PDF
  (the transformation brings in a Jacobian);
  it has to be adjusted with care.

DFM: SHOULD WE DO THIS ALL EXPLICITLY HERE IN MATH?

@@

%% \section{Advice and discussion}

@@

%% ...Sample the prior (likelihood set to unity) to check priors and parameterization (another functional test).

%% ...Posteriors: On the other hand, you can do anything you like (within
%% reason) to the output samples and be sampling the posterior in that
%% transformed quantity.

>>>>

Say you are in this situation.
You have a huge blob of data $\data$ (think of this as a vector or
list or heterogeneous collection of observations\note{In general in
  this \documentname\ we will treat any variable as being capable of
  containing not just a value but some huge arbitrarily structured set
  of values in matrix, vector, or more complex form.}).
You also have a model sophisticated enough---a probabilistic,
generative model, if you will\note{For our definition of a
  \emph{model}, see [SOME OTHER NOTE] in this series.}---that, given a
setting of a huge blob of parameters\note{Again, the parameter
  blob $\pars$ should also be thought of as an immense, arbitrarily
  structured set or list of parameters, possibly even with no fixed
  dimension!} $\pars$, you can compute a pdf for data (or likelihood\note{We
  bash the terminology ``likelihood'' in [SOME OTHER NOTE] in this
  series.}) $p(\data\given\pars)$.
% ************* BJB
% <Wild Subjective Bayesian>
% The "generative model" is a probabilistic description of your *prior
% beliefs about what the data are going to be*. </Wild Subjective Bayesian>
% ************* Hogg
% That comment should go in the ``What is a Model?'' Chapter.
% (Or a \note{} right here.)
Furthermore, say you also can write down some kind of informative or
vague prior pdf $p(\pars)$ for the parameter blob.
If all these things are true, then---even if you can't compute
anything else---in principle a trivial-to-implement MCMC can give you
a fair sampling of the posterior pdf.
That is, you can run MCMC (for a very long time\note{Sometimes a very,
  very long time; we will return to that below.})  and you will be
left with a set of $K$ parameter-blob settings $\pars_k$ such that the
full set $\setof{\pars_k}_{k=1}^K$ constitutes a fair sampling from
the posterior pdf $p(\pars\given\data$).
% ************* BJB
% Notions of "sampling" and "Monte Carlo" are interesting. You might like to
% refer the interested reader to the paper "Monte Carlo is Fundamentally
% Unsound" by Tony O'Hagan.

@@

Imagine that you are in the simplest case:
You have a model with a small number of parameters (say three-ish),
  and only one of them is of great interest.
What are your options for the ``measurement'' of this parameter?
The only simple integral-based options are the posterior mean
  or posterior median value for the parameter.

For example,
  imagine that by MCMC you have generated
  $K$ samples $\pars_k$ of a parameter vector $\pars$.
Now you have a scalar function $q(\theta)$ which takes the parameter vector
  and returns a scalar value%
  \note{For the median-of-sampling value, $q(\theta)$ needs to return a scalar,
  but technically, for the mean-of-sampling, $q(\theta)$ can be a vector or something high-dimensional.}
  which could be as simple as a single component of the parameter vector
  (one parameter from the list),
  or something more complex.
The mean-of-sampling value $\mean{q}$ is just
\begin{eqnarray}
\mean{q} &\leftarrow& \frac{1}{K}\,\sum_{k=1}^K q(\theta_k)
\quad ,
\end{eqnarray}
  and the median-of-sampling value is just the $[K/2]$th value of $q(\theta_k)$
  when the $q(\theta_k)$ have been ordered (sorted).
These are both produced by integrals;
  the first is the value returned by the expectation integral estimate,
  the second is the value past which half of the integrated pdf lies.

What are your best options for the ``uncertainty'' or ``limits'' on this parameter?
The only simple integral-based options are posterior quantiles.
That is, the ``one-sigma'' error bar can be the size of the central (or smallest%
  \note{HOGG shizzle about choosing quantiles}) interval of the parameter
  that contains 68 percent of the posterior samples.
The ``two-sigma'' error bar would be the same but for the 95-percent interval.
Again, these limits can be estimated%
  \note{One amusing thing about samplings,
  which are so beloved of us probabilistic (Bayesian) reasoners,
  is that anything we \emph{do} with a sampling,
  like estimate an integral or a quantile,
  is just an old-school frequentist estimator.
  A frequentist estimator of a Bayesian quantity, to be sure, but a frequentist estimator nonetheless.}
  by ordering the $q(\theta_k)$ values,
  and finding the $\theta$ values such that some fraction (0.68 or 0.95)y
  of the samples lie between them.
Again, to form these limits, $q(\theta)$ must be a scalar function (so sorting is well defined).

One amusing thing is that if you choose the 68-percent interval sensibly,
  but use as the ``measurement'' the posterior mean,
  you can get pathological situations in which the posterior mean measurement
  is actually \emph{outside} the one-sigma confidence interval.
For this reason (and others), we usually recommend as a default behavior%
  ---in the one-dimensional case---%
  to choose the \emph{median} of sampling as the measurement value,
  the 16-percent quantile as the lower one-sigma error bar,
  and the 84-percent quantile as the upper one-sigma error bar.
This has pathologies in higher dimensions (as we are about to see),
  but is pretty safe for one-dimensional answers.

@@

Another proxy for autocorrelation time useful for tuning is the
  Expected Squared Jump Distance.\note{(CITE GELMAN)}
This is the mean squared distance the walker moves, per step.
It is maximized when the acceptance ratio is reasonable
  and the step size is large;
  it is large when the exploration of the space is fast.
This proxy is easy to measure and use for tuning,
  and more directly related to autocorrelation time than the acceptance ratio.
We recommend using it for tuning, though we have never used it ourselves.
Like the acceptance fraction, it can also be used to tune
  in the case that the proposal distribution loops over parameters;
  once again, the user gets (in this case) $d$ one-dimensional tunings.

@@

MCMC is used to obtain samples $x_k$ from pdf $p(x)$,
  or a badly normalized all-positive function $f(x)$
  that is different from the pdf by an unknown factor $Z$.
In the context of data analysis,
  MCMC is usually being used to obtain samples
  that are \emph{parameter values} for a probabilistic model of some data.
That is, MCMC is sampling a pdf \emph{for the parameters}.
If it is returning parameter values, then it is not sampling a pdf for the data.
That is, MCMC can only sample a posterior (or prior) pdf for the parameters.
\emph{It cannot sample a likelihood.}%
\note{Well, technically, MCMC \emph{can} be used to sample a likelihood function,
  but the samples would be samples of possible \emph{data} not possible \emph{parameters}.
  The purist might say that even this is not sampling a likelihood function,
  because you should only call it a ``likelihood function'' in contexts in which you are
  treating the data as fixed (at the true data values) and the parameters as variable.
  In this context, the likelihood function is \emph{not} a pdf for the parameters,
  so it can't be sampled.}

@@

\paragraph{parameterization problems:}
It helps to think carefully about how to parameterize.
For example, if there is an angle $\phi$ in your problem,
  and your prior requires it to be between $0$ and $2\pi$,
  and the posterior mode is very near $0$,
  a little bit of probability leaking below $0$ plus the prior cutoffs plus mod-$2\pi$ symmetry
  makes an intrinsically unimodal problem multimodal!
In these cases, we usually advise reparameterization from a (say) amplitude $A$ and angle $\phi$
  to two vector components $a\equiv A\,\cos\phi$ and $b\equiv A\,\sin\phi$.
Of course this re-parameterization changes significantly the \emph{form} of the prior PDF
  (the transformation brings in a Jacobian);
  it has to be adjusted with care.

DFM: SHOULD WE DO THIS ALL EXPLICITLY HERE IN MATH?
Choose a $p(A)$ and a $p(\phi) = [2\pi]^{-1}$,
   and then transform to $p(a, b)$.

@@

@@

%% ...Posteriors:  You can do anything you like (within
%% reason) to the output samples and be sampling the posterior in that
%% transformed quantity.
